\appendix

\chapter{Datasets in Detail}
\label{app:datasets}

This appendix consolidates the per-source / per-clip dataset metadata that the methodology chapter summarises. Numbers come from the dataset YAMLs (\texttt{rgb\_dataset/dataset.yaml}, \texttt{IR\_dset\_final/dataset.yaml}), the documentation file shipped with the confuser corpus (\texttt{rgb\_confusers\_merged/dataset\_documentation.md}), the Svanstr\"om paired-corpus meta file, and the YouTube extraction manifests.

\section{Composite RGB Training Corpus}

The full distribution by source prefix is given in Table~\ref{tab:ds_rgb_components}; the split policy is 80 / 10 / 10 with sequence-disjoint partitioning where the source did not already supply its own splits. Six negative-source rows (VIRAT, UA-DETRAC, BDD100K) contribute the 21.9\% background-negative fraction. The AirBird subset (10{,}000 images) is the bird hard-negative channel the baseline detector relies on; the composite corpus does not contain Svanstr\"om bird / airplane / helicopter splits, which are therefore OOD at evaluation time.
% [source: dataset=rgb_dataset; file=rgb_dataset/dataset.yaml]

\section{Confuser Hard-Negative Corpus}

Per-source composition is in Table~\ref{tab:ds_confusers}. Split policy:

\begin{itemize}
    \item \textbf{Sources with existing splits} (Airplane Roboflow, New\_Dataset Roboflow): original train / valid / test partitions preserved.
    \item \textbf{Sources without splits}: split by video / sequence ID at 80 / 10 / 10 to prevent leakage.
    \begin{itemize}
        \item Svanstr\"om: 165 video sequences split by sequence ID.
        \item raihanrsd: 58 video folders split by folder.
        \item Helicopter Kaggle: 20 hash-based pseudo-groups.
    \end{itemize}
    \item \textbf{Mislabeled frame removal}: 10 frames in the airplane / New\_Dataset sources contained visible drones and were excised before training.
\end{itemize}

\noindent The mined-hard-negative process for \texttt{retrained\_v2} ran the baseline detector on every confuser image and kept the 11{,}729 frames (43.4\%) on which the baseline fired; those mined frames were merged with the composite RGB corpus to produce the final 183{,}751-image \texttt{retrain\_dataset}. \texttt{hardneg\_v3more} differs in using a smaller, less aggressive merge.
% [source: dataset=rgb_confusers_merged; file=rgb_confusers_merged/dataset_documentation.md]

\section{IR Detector Corpus}

The per-prefix distribution is in Table~\ref{tab:ds_ir_components}. The Svanstr\"om IR partition (\texttt{svan}, 21{,}637 frames across the corpus' train/val/test splits: 17{,}314 train, 2{,}050 val, 2{,}273 test, plus 1{,}325 zoom-augmented copies under \texttt{czoom}) is the largest single Svanstr\"om appearance in any training corpus in this thesis and is the reason Svanstr\"om IR evaluation is in-distribution for the IR detector (Table~\ref{tab:svanstrom_audit}).
% [source: ls G:/drone/IR_dset_final/{train,val,test}/images svan_* counts 17314/2050/2273; czoom_svan_* 1325; thesis_eval/results/leakage_controlled.json derived] The corpus state documented here is the state at the \texttt{Final} checkpoint; the V2--V6 revision history (Section~\ref{sec:ir_evolution}) operated on smaller predecessor variants of the same source families.

\section{Thermal Confuser Corpus (\texttt{IR\_confusers})}

5{,}938 thermal frames containing aerial confusers and no drones: 4{,}281 airplane, 1{,}200 bird, and 457 helicopter frames (category encoded in the filename prefix), exported from Roboflow-hosted thermal confuser video; source streams disjoint from every \texttt{ir\_dset\_final} source family. The evaluated split holds 5{,}237 frames, sampled to 4{,}000 by even spread (3{,}043 airplane / 871 bird / 86 helicopter). All frames carry empty YOLO labels; every detection on this surface is a false positive by construction. The corpus contributes no training data to any model; it is the OOD stress surface for the IR branch (Section~\ref{sec:ds_ir_confusers}). The thermal-native head is trained with the GT-aware CBAM \emph{train} split; its held-out gates are the disjoint \emph{CBAM valid} split (drone recall recovered $0.717 \to 0.967$) and the held-out \texttt{IR\_confusers} val/test re-mine ($90 \to 22$ residual fires, $94\%$ removed; Section~\ref{sec:ir_xmodal_verifier}).

\section{Svanstr\"om Paired}

28{,}710 frames, sampled at stride~3 from 279 source video sequences. Per-class instance counts: drone 11{,}695; airplane 6{,}090; bird 5{,}298; helicopter 5{,}627. Native resolution $640{\times}512$ (RGB) with a co-registered IR channel at the same resolution. Annotators drew loose boxes around small drones, motivating the IoP@0.5 scoring rule of Section~\ref{sec:metrics}. The dataset ships as source videos with MATLAB \texttt{groundTruth} annotation objects~\cite{svanstrom2022dronedataset}; the frames and YOLO labels used here were produced by this project's own conversion script, so box geometry may differ marginally from the authors' MATLAB pipeline, and the original videos are retained alongside the converted corpus.
% [source: github.com/DroneDetectionThesis/Drone-detection-dataset; conversion = project script; originals at G:/drone/svanstrom_paired]

\section{Anti-UAV RGBT}

The converted YOLO test + val splits used for evaluation comprise 85{,}374 paired frames at $1920{\times}1080$ (RGB) and $640{\times}512$ (IR). Anti-UAV is a tracking benchmark in its native form; the per-frame detection evaluation reported in this thesis is computed by re-scoring the tracking benchmark's frame-level GT boxes with the IoU@0.5 matching rule.

\section{DUT Anti-UAV}

The DUT Anti-UAV detection dataset~\cite{zhao2022dutantiuav} ships $10{,}000$ aerial RGB images with author-provided train / validation / test partitions. It enters this thesis twice. As a \emph{training} source, $5{,}200$ of its images contribute to the composite RGB corpus (the \texttt{dut} prefix in Table~\ref{tab:ds_rgb_components}), so the production RGB detector \texttt{ft4} is in-distribution on it. As an \emph{evaluation} surface, the official test split (the $2{,}200$ frames / $2{,}245$ ground-truth drones used in Table~\ref{tab:ablation_dut}) is the confuser-free, all-drone RGB check of Section~\ref{sec:pipeline_paired}, evaluated at \texttt{imgsz=960} with the thermal detector in its grayscale-fallback mode. No DUT frame used at evaluation time was a training image for the trust router or the per-frame filters.
% [source: github.com/wangdongdut/DUT-Anti-UAV; cite=zhao2022dutantiuav; eval split = thesis_eval/results_dut/tier1_results.json]

\section{SelCom CCTV}

Single-camera source: fixed perimeter installation, $1920{\times}1080$, 25~fps. Source clip $\sim$1:30, 2{,}076 extracted frames (1{,}953 positive + 123 explicit negative). Median drone $\sqrt{\text{area}} \approx 36.8$~px in the native frame (so $\sim$12~px at \texttt{imgsz=640}, $\sim$24~px at \texttt{imgsz=1280}). The 80 / 20 train / val split (20\% pure-SelCom validation, seed~0) is preserved across the \texttt{ft1} / \texttt{mixed\_ft1} / \texttt{mixed\_ft2} / \texttt{mixed\_ft2\_1280} variants; the strongest CCTV-evaluated fine-tune is \texttt{Yolo26n\_selcom\_mixed\_ft2\_1280}, a comparison detector; the production RGB detector is \texttt{ft4} (Section~\ref{sec:production_stack}).
% [source: dataset=selcom_cctv; file=SelCom CCTV manifest (see Section~\ref{sec:ds_selcom})]

\section{YouTube Real-Video Diagnostic}

\begin{table}[htbp]
    \centering
    \caption{Drone-positive YouTube clips. ``Extracted'' is the frame count used in evaluation after stride sampling; total native frame count is shown for context. ``Source'' is the YouTube video ID (prepend \texttt{https://youtu.be/}), recovered from the extraction manifest.}
    \label{tab:youtube_drone_clips}
    \footnotesize
    \begin{tabular}{lrrrrl}
        \toprule
        Clip & FPS & Total frames & Stride & Extracted & Source \\
        \midrule
        \texttt{drone\_and\_bird\_sky\_and\_trees\_short}     & 30.0 &   506 & 0 (full) & 114 & \texttt{KoUgV6WhL8g} \\
        \texttt{drone\_attacked\_by\_bird\_mountain\_side\_view} & 30.0 & 400 & 3 & 133 & \texttt{yAbMFQrt98M} \\
        \texttt{drone\_over\_mountain\_attacked\_by\_birds}   & 30.0 &   325 & 3 & 108 & \texttt{1qRERzVpRFE} \\
        \texttt{drone\_seagull\_attack}                        & 30.0 & 1{,}146 & 0 (full) & 235 & \texttt{nTdbEPxlqPM} \\
        \texttt{drone\_takeoff\_from\_ground\_and\_not\_hand\_short} & 60.0 & 1{,}379 & 0 (full) & 163 & \texttt{85XdkfWsCuI} \\
        \texttt{drone\_takeoff\_short}                         & 60.0 & 637 & 0 (full) & 116 & \texttt{Qx1Hlot9Ye8} \\
        \texttt{drone\_takeoff\_short\_trees\_background\_dji} & 60.0 & 954 & 0 (full) & 166 & \texttt{J-gcjDUjg0A} \\
        \texttt{flock\_of\_birds\_attack\_drone}              & 30.0 & 338 & 3 & 112 & \texttt{OAjZ2bgh-Xc} \\
        \texttt{flock\_of\_seagulls\_attack\_drone\_beach}    & 60.0 & 1{,}010 & 3 & 336 & \texttt{DdkFYVsXAUQ} \\
        \texttt{two\_birds\_drone}                              & 30.0 & 456 & 0 (full) & 150 & \texttt{ngMjVZfdOc\_} \\
        \midrule
        \textbf{Total} & & 7{,}151 & & \textbf{1{,}633} & \\
        \bottomrule
    \end{tabular}
    \normalsize
\end{table}
% [source: datasets/drone detection video tests/rgb/extraction_manifest.json (IDs from the _Media_ filename marker)]

\noindent Note: the evaluated drone-positive set is the 9-clip / 1{,}359-frame / 1{,}234-GT-instance subset of this table that re-ran cleanly through the corrected-scoring pipeline (\texttt{flock\_of\_birds\_attack\_drone} excluded; some clips hold slightly fewer frames on disk than at extraction time). All Chapter~\ref{ch:hitl} video numbers, including the temporal evaluation, use that evaluated subset.

\begin{table}[htbp]
    \centering
    \caption{Confuser-only YouTube clips. All frames are confuser-positive only; every detection in these clips is a false positive by construction. ``Source'' is the YouTube video ID (prepend \texttt{https://youtu.be/}); two compilation clips were saved without the ID marker.}
    \label{tab:youtube_confuser_clips}
    \footnotesize
    \begin{tabular}{lrrrrl}
        \toprule
        Clip & FPS & Total frames & Stride & Extracted & Source \\
        \midrule
        \texttt{airplanes\_airplanes\_compilation}             & 25.0 & 6{,}248 & 25 & 249 & not preserved \\
        \texttt{airplanes\_distant\_airplane\_over\_head\_flying\_away} & 29.0 & 1{,}606 & 29 & 55 & \texttt{ANbApycXbJ4} \\
        \texttt{birds\_birds\_flying\_overhead\_various\_sizes\_short} & 30.0 & 201 & 10 & 20 & \texttt{WytG-t1S6G0} \\
        \texttt{birds\_birds\_in\_slow\_motion\_flying\_various\_sizes\_compilation} & 50.0 & 13{,}550 & 50 & 271 & \texttt{ZO5lV0gh5i4} \\
        \texttt{birds\_distant\_birds\_flying\_in\_the\_sky\_short} & 30.0 & 462 & 23 & 20 & \texttt{4zMELhui9DE} \\
        \texttt{birds\_flock\_of\_birds\_flying\_short}        & 30.0 & 259 & 12 & 21 & \texttt{vmHsvO83wrk} \\
        \texttt{birds\_flock\_of\_birds\_flying\_sunset}      & 24.0 & 492 & 24 & 20 & \texttt{q0mmJrkQYaU} \\
        \texttt{helicopters\_helicopter\_compilation}          & 25.0 & 13{,}871 & 25 & 554 & not preserved \\
        \texttt{helicopters\_helicopter\_overhead\_short}      & 25.0 & 260 & 13 & 20 & \texttt{xp8oiz8ymbU} \\
        \texttt{helicopters\_helicopter\_overhead\_very\_small\_airplane\_in\_background} & --- & --- & --- & 20 & \texttt{7mPDG7wrH5I} \\
        \midrule
        \textbf{Total} & & 36{,}949 & & \textbf{1{,}250} & \\
        \bottomrule
    \end{tabular}
    \normalsize
\end{table}
% [source: datasets/confuser_videos/extraction_manifest.json (IDs from the _Media_ filename marker)]

\noindent The 10 confuser clips were drawn from public YouTube sources visually selected to span the three confuser categories at multiple ranges and viewing angles; the video IDs above are recovered from the \texttt{\_Media\_} marker the downloader embedded in the clip filenames, preserved in \texttt{datasets/confuser\_videos/extraction\_manifest.json}. The aggregate confuser-FPPI numbers in Section~\ref{ch:experiments} use the 1{,}250-frame totals.

\section{Licensing and Ethics}

The public benchmark datasets (Anti-UAV, Svanstr\"om, the Roboflow community sets) are used under their published research licenses; the IR component sources (FLIR, DV5, CST, etc.) are similarly research-license open. The SelCom CCTV stream is deployment-partner proprietary and is not redistributable; only summary metrics and qualitative figures derived from it appear in this thesis. The YouTube clips are short, low-resolution excerpts used under fair-use for research and contain no personally identifying information; the encoded source IDs in the clip filenames are preserved so the originals can be located by an interested party. None of the corpora include human faces or location markers beyond what is visible in the original public datasets.

% ======================================================================
% APPENDIX B — GLOSSARY
% ======================================================================
% === BEGIN app:models (generated by docs/gen_models_appendix.py) ===
\chapter{Models Evaluated}
\label{app:models}

This appendix lists the models the thesis actually discusses: the production pick at each cascade stage (\textbf{production}) and the comparison or ablation variants that carry a finding in the main text. The complete experimental registry, including every intermediate, superseded, and archived checkpoint, is retained in the project repository for reproducibility. The IR detector's full six-revision evolution is reported separately in Table~\ref{tab:ir_evolution}.

\footnotesize
\begin{longtable}{p{0.27\textwidth}lp{0.53\textwidth}}
\toprule
Model & Role & Note \\
\midrule
\endhead
\multicolumn{3}{l}{\textbf{RGB detectors (YOLOv26n)}} \\
\midrule
\texttt{ft4} & \textbf{production} & SelCom-plus-confuser fine-tune; the only confuser-injection recipe that passed every drone-recall regression gate. \\
\texttt{baseline} & comparison & Stage-1 reference; best small-drone recall at \texttt{imgsz=1280}; base for the SelCom fine-tunes. \\
\texttt{hardneg\_v3more} & comparison & Hard-negative mining: lowers helicopter and airplane fire but leaves bird fire essentially unchanged. \\
\texttt{retrained\_v2} & comparison & Aggressive confuser retrain; suppresses birds but collapses Svanstr\"om drone recall to $R=0.306$. \\
\texttt{selcom\_mixed\_ft2\_1280} & comparison & CCTV (SelCom) fine-tune weights at \texttt{imgsz=1280}. \\
\midrule
\multicolumn{3}{l}{\textbf{IR detectors (YOLOv26n, thermal)}} \\
\midrule
\texttt{ir\_v3b} & \textbf{production} & Production IR detector; $F1=0.967$ on the fixed IR test split; two-epoch corrective fine-tune of \texttt{Final}. \\
\texttt{Final} & comparison & Immediate parent of \texttt{v3b}; the full V2--V6 / \texttt{Final} trajectory is in Table~\ref{tab:ir_evolution}. \\
\midrule
\multicolumn{3}{l}{\textbf{Trust classifiers (XGBoost fusion / routing)}} \\
\midrule
\texttt{robust8-nr} & \textbf{production} & Shipped trust router: the eight \texttt{robust8} features with the \texttt{reject} class dropped (always routes RGB/IR/both; the filter owns false-positive rejection). \\
\texttt{robust8} & ablation & Four-class router (\texttt{robust6} $+$ two free features $+$ a tuned per-class threshold); recommended on confuser-rich paired streams, retained as the reject-class ablation. \\
\texttt{robust6} & comparison & Six statistically-selected, leakage-free features; the base extended into \texttt{robust8}/\texttt{robust8-nr}. \\
\texttt{sa32} & comparison & Strongest hand-engineered (32-feature) classifier; best in-domain on Svanstr\"om, but leaks scene statistics so its lead is not trusted to generalise to unseen surfaces; superseded in production by \texttt{robust8-nr}. \\
\texttt{control40} & comparison & 40-feature control: best raw drone metrics but the worst confuser rejection. \\
\texttt{fusion\_no\_fn\_v1.1} & comparison & Open-world fallback; the most conservative variant on the OOD confuser zoo. \\
\midrule
\multicolumn{3}{l}{\textbf{Confuser filters}} \\
\midrule
\texttt{mlp\_v5} (\texttt{v4} build) & \textbf{production} & Production RGB filter: a distilled MLP on the detector's own \texttt{p3}+\texttt{p5} ROI features, run per-frame; the \texttt{v4} bird-split build closes the \texttt{rgb\_dataset\_test} coverage gap ($F1\;0.809\to0.922$, recall $0.691\to0.887$). \\
\texttt{mlp\_aligned\_thermalonly} & \textbf{production} & Thermal-native head of the two-head IR filter (CBAM-trained, @0.05); held-out CBAM recall recovered $0.717\to0.967$ at FP $6$, with $94\%$ of held-out thermal confusers removed. \\
\texttt{mlp\_aligned\_gray\_balanced} & \textbf{production} & Grayscale-aligned head of the two-head IR filter (@0.25); cuts held-out grayscale confuser FPs by ${\sim}97.7\%$ ($656\to15$). \\
\texttt{patch\_v2} & comparison & Superseded MobileNetV3 patch filter; audited predecessor, the \texttt{v4} RGB filter removes its last fallback role. \\
\bottomrule
\caption{Models referenced in this thesis: the production pick at each cascade stage and the comparison or ablation variants discussed in the main text. The complete experimental registry is retained in the project repository.}
\label{tab:models_evaluated}
\end{longtable}
\normalsize

% === END app:models ===

\chapter{Model MRI Sample Report}
\label{app:mri_report}

The Model MRI renders a human-readable report for every run (Section~\ref{sec:model_mri}). Reproduced below, lightly typeset but textually verbatim, is the report for the production RGB filter's corpus (\texttt{mri/results/v5\_report\_regen/report.md}); the excerpt in Figure~\ref{fig:mri_report} is its verdict block.

\begin{quote}\small\ttfamily
\textbf{Model MRI --- Yolo26n\_selcom\_confuser\_ft4\_1280}\\[2pt]
Detector: \texttt{RGB model/Yolo26n\_selcom\_confuser\_ft4\_1280/weights/best.pt}\\
Corpus: resumed feature corpus --- 19{,}334 drone / 13{,}597 confuser detections\\[6pt]
\textbf{Verdict:} Classifier strongly recommended --- large FP cut at low recall cost.\\[2pt]
Evidence: 19{,}334 drones / 13{,}597 confusers; LDA separability 0.952 (thr 0.90); recall cost 1.1\% (thr 10\%); projected FP cut 97\%.\\[6pt]
\textbf{Diagnostic signals}\\[2pt]
\begin{tabular}{@{}lll@{}}
Signal & Value & Meaning \\
\midrule
Raw hallucination rate & 54.4\% & FP per confuser image (bare FT4 on \texttt{rgb\_confusers} @1280) \\
LDA separability & 0.952 & train-set linear split of drone vs confuser \\
Silhouette & 0.067 & feature-space cluster separation (unsupervised) \\
Max ANOVA $F$ & 42{,}346 & strongest single discriminative feature \\
Meta-only max AUROC & 0.811 & best metadata feature alone \\
YOLO-feat max AUROC & 0.844 & best learned feature alone \\
Projected FP cut & 97.4\% & confusers the classifier would reject \\
Recall retention & 98.9\% & true drones the classifier keeps \\
Projected FP rate & 1.4\% & hallucination after classifier \\
\end{tabular}\\[6pt]
\textbf{Top discriminative features (ANOVA $F$):} p5 ch=154; p3 ch=31 cell=0; p5 ch=79; p3 ch=31 cell=1; p5 ch=113; meta:conf; p5 ch=240; p5 ch=85; p3 ch=31 cell=2; p5 ch=43.
\end{quote}

The report's projection (97.4\% FP cut at 98.9\% recall retention) is the pre-training estimate; the shipped filter's measured behaviour on the canonical surfaces is in Section~\ref{sec:verifier_results} and Table~\ref{tab:ablation_confusers}. The agreement between projection and measurement is itself part of the instrument's validation.

\chapter{Number Provenance and Reproducibility}
\label{app:provenance}

Every headline number in this thesis is the value of a named cell in a canonical results file, and every such file is reproducible by a stated command. The map lives in the project repository as \texttt{runs/README.md}; this appendix reproduces it. Three layers make the chain auditable:

\begin{enumerate}
    \item \textbf{Unified detection cache.} \texttt{thesis\_eval/pipeline\_cache\_unified.py} runs the production detectors \emph{once} per evaluation surface and stores, per frame, every detection (box, confidence, the 517-D ROI feature vector, patch-classifier probability), the trust classifier's input features, and the per-modality ground truth. Caches are tagged with the detector weight identity.
    \item \textbf{Deterministic replays.} Every evaluation in Chapter~\ref{ch:hitl} is a CPU re-score of that cache (no detector forward pass), so any reader with the repository can re-derive any table in about a minute per surface. The replay scripts and their outputs are listed in Table~\ref{tab:provenance}.
    \item \textbf{Automated audit.} \texttt{thesis\_eval/\_audit\_headline\_numbers.py} asserts, cell by cell, that the numbers printed in this document equal the values in the canonical results files; the submission build of this thesis passes that audit.
\end{enumerate}

\begin{table}[htbp]
    \centering
    \caption{Headline-number provenance: thesis claim $\to$ canonical results file (under \texttt{runs/}, frozen copies of \texttt{thesis\_eval/results/}) $\to$ reproduction command. All \emph{pipeline} replay commands are zero-GPU; the two rows marked \texttt{(GPU)} (the IR-detector trajectory and stage latencies) require the detectors and target hardware.}
    \label{tab:provenance}
    \footnotesize
    \begin{tabular}{p{4.4cm}p{4.6cm}p{4.6cm}}
        \toprule
        Thesis number & File (cell) & Reproduce \\
        \midrule
        Svanstr\"om composed F1 $0.742 \to 0.946$ (shipped \texttt{robust8-nr}, \texttt{filt$\to$clf}) & \texttt{results\_noreject/tier1\_results.json} (\texttt{svanstrom.B\_pipeline}) & \texttt{pipeline\_eval\_unified.py --only svanstrom} \\
        Confuser fire $30.4\% \to 1.4\%$ (shipped \texttt{robust8-nr}; $0.11\%$ \texttt{robust8} ablation) & \texttt{results\_noreject/tier1\_results.json} (\texttt{rgb\_confuser.C\_confuser}) & \texttt{pipeline\_eval\_unified.py --only rgb\_confuser} \\
        Anti-UAV no-harm $0.973 \to 0.984$ & \texttt{tier1\_results.json} (\texttt{antiuav.B\_pipeline}) & \texttt{pipeline\_eval\_unified.py --only antiuav} \\
        Grayscale 3-way $0.607/0.187/0.580$ & \texttt{tier1\_results.json} (D table) & \texttt{pipeline\_eval\_unified.py} \\
        Temporal 2-of-3 windows (Table~\ref{tab:temporal_production}) & \texttt{temporal\_results.json} & \texttt{temporal\_replay.py} \\
        Per-size buckets (Table~\ref{tab:per_size}) & \texttt{notes\_round1\_results.json} (\texttt{SZ\_per\_size}) & \texttt{notes\_round1\_replays.py} \\
        Background profile (Table~\ref{tab:failure_profile}) & \texttt{failure\_profile\_results.json} & \texttt{failure\_profile\_aggregate.py} \\
        Low-conf mode (Table~\ref{tab:lowconf_selcom}) & \texttt{conf\_sweep/conf\_sweep\_results.json} & \texttt{conf\_sweep\_replay.py} \\
        IR HITL trajectory $0.503 \to 0.967$ & \texttt{knowledge/evals.csv} (\texttt{ir\_final\_*}, one fixed split) & \texttt{eval/ir\_version\_comparison.py} (GPU) \\
        Stage latencies (Table~\ref{tab:speed}) & \texttt{knowledge/ledger.csv} (speed rows) & \texttt{eval/bench\_speed.py} (GPU) \\
        Scoring swing $2.8$~pp (Section~\ref{sec:scoring_audit}) & \texttt{notes\_round1\_results.json} vs \texttt{tier1\_results.json} (routed \texttt{+filt} / \texttt{clf->filt}) & \texttt{notes\_round1\_replays.py} \\
        Overlap audit counts (Table~\ref{tab:svanstrom_audit}) & \texttt{clean\_split/README.md} $+$ \texttt{leakage\_controlled.json} & \texttt{leakage\_controlled\_replay.py} \\
        Clean-split results (Table~\ref{tab:clean_split}) & \texttt{clean\_split/clean\_split\_results.json} (sequence lists alongside) & cache \texttt{--only svanstrom\_clean,antiuav\_clean --full}, then \texttt{pipeline\_eval\_unified.py} \\
        \bottomrule
    \end{tabular}
\end{table}

The \texttt{runs/} directory additionally holds the production IR detector's complete Ultralytics training record (\texttt{runs/corrective\_finetune/}: training arguments, loss and PR curves, confusion matrices, and acceptance reports for the \texttt{v3b} fine-tune), so the training side of the production stack is documented alongside the evaluation side. The relational evidence store \texttt{knowledge/} (CSV tables with generated views) records every evaluation and finding as a row carrying its source path and reproduction command.
% [source: runs/README.md (this appendix mirrors it)]

\chapter{Glossary of Abbreviations and Shorthand}
\label{app:glossary}

\begin{description}
    \item[C-UAS] Counter-Unmanned Aerial Systems. The defensive use-case context for this thesis.
    \item[UAV] Unmanned Aerial Vehicle, used interchangeably with ``drone'' for consumer-class quadcopters in this thesis.
    \item[YOLO] You Only Look Once. The single-stage object-detection family~\cite{redmon2016yolo}; this thesis uses YOLOv26n via the Ultralytics framework~\cite{ultralytics2024}.
    \item[CNN] Convolutional Neural Network. The patch filter is a MobileNetV3-class CNN~\cite{howard2019mobilenetv3}.
    \item[HITL] Human-in-the-Loop. The dataset / model co-development protocol of Section~\ref{sec:hitl_ir}.
    \item[OOD] Out-Of-Distribution. Imagery drawn from a different population than the model's training distribution.
    \item[FPPI] False Positives Per Image. The per-frame false-alarm count, used as the primary metric on confuser-only evaluation surfaces.
    \item[FPR] False Positive Rate. In this thesis: the frame-level fraction of confuser-only frames with at least one detection.
    \item[IoU] Intersection over Union. The standard detection-matching rule; used for Anti-UAV scoring.
    \item[IoP] Intersection over Prediction. The matching rule used for Svanstr\"om RGB scoring (Section~\ref{sec:metrics}); $\text{IoP} = |B_\text{pred} \cap B_\text{gt}| / |B_\text{pred}|$.
    \item[S1 / S2 / S3] Cascade stages: S1 = RGB detector alone; S2 = + trust classifier; S3 = + patch filter alert gate.
    \item[ft4] \texttt{Yolo26n\_selcom\_confuser\_ft4\_1280}. The production RGB detector (SelCom$+$confuser fine-tune); the only confuser-injection config that passed every drone-recall regression gate (Section~\ref{sec:training_recipes}).
    \item[robust6] The six-feature trust classifier (per-modality detection confidence $+$ box geometry), its features screened by the scene-fingerprint ratio (Section~\ref{sec:feature_selection}); the base that \texttt{robust8} extends.
    \item[robust8] The four-class trust classifier (Section~\ref{sec:feature_selection}): \texttt{robust6} plus two further free features and a tuned per-class decision threshold; retained as the reject-class ablation and recommended on confuser-rich paired streams.
    \item[robust8-nr] The shipped production trust router (Section~\ref{sec:classifier_results}): the eight \texttt{robust8} features with the \texttt{reject} class dropped, so it always routes RGB/IR/both and the per-frame filter owns false-positive rejection.
    \item[sa32] \texttt{scene\_aware\_v3more\_32feat}. The strongest hand-engineered comparison trust classifier (32-feature XGBoost); it leaks scene statistics (discriminates \emph{which scene} a frame came from, not \emph{whether} its detections are trustworthy), so its in-domain lead is not trusted to generalise, and it is superseded in production by \texttt{robust8-nr} (Section~\ref{sec:feature_selection}).
    \item[fnfn] \texttt{fusion\_no\_fn\_v1.1}. The open-world fallback trust classifier (40-feature).
    \item[control40] \texttt{control\_v3more\_40feat}. A previous-candidate 40-feature trust classifier; deprecated.
    \item[v3b] The production IR detector weights: 2-epoch corrective fine-tune on top of the \texttt{Final} IR checkpoint.
    \item[v2 (patch filter)] \texttt{confuser\_filter4\_\{rgb,ir\}\_v2\_backup.pt}. The MobileNetV3 patch filter; the audited predecessor, superseded in production by \texttt{mlp\_v5}.
    \item[\texttt{mlp\_v5}] The distilled feature-space confuser filter (Section~\ref{sec:distill_verifier}): an MLP on the detector's fused \texttt{p3}+\texttt{p5} ROI features that supersedes the v2 patch filter on confuser-rich surfaces and runs per-frame; production is the \texttt{v4} bird-split build.
    \item[IR filter (two heads)] The cross-modal IR confuser filter (Section~\ref{sec:ir_xmodal_verifier}), now two heads selected per modality: a thermal-native head (\texttt{mlp\_aligned\_thermalonly}, CBAM-trained, @0.05) for thermal deployment and a grayscale-aligned head (\texttt{mlp\_aligned\_gray\_balanced}, @0.25) for the grayscale-fallback path.
    \item[selcom\_1280] The SelCom CCTV RGB fine-tune \texttt{Yolo26n\_selcom\_mixed\_ft2\_1280} evaluated at \texttt{imgsz=1280}; a comparison/ablation detector (the production RGB detector is \texttt{ft4}, Section~\ref{sec:production_stack}).
    \item[selcom\_640] The same SelCom weights evaluated at \texttt{imgsz=640} (resolution-ablation variant).
    \item[\texttt{imgsz}] Ultralytics short for input letterbox size. Drives small-object recall (Section~\ref{sec:resolution_arch}).
    \item[\texttt{patch\_thr}] The decision threshold on $p_\text{confuser}$ at the patch-filter alert gate. Production: 0.9 on Svanstr\"om-like inputs, 0.7 on real-video.
\end{description}
