\chapter{Conclusion}
\label{ch:conclusion}

This thesis set out to make visual drone detection deployable: high recall and a low false-alarm rate at the same time, on hardware a site can afford. The position it defends is architectural: the base detectors are tuned purely for recall, and every discrimination they cannot learn without losing recall (confuser rejection, modality arbitration, false-positive filtering) is relocated into cheap, learned, downstream stages. The evaluation holds one protocol throughout (Section~\ref{sec:eval_protocol}); every number below carries its configuration and its confidence interval in the tables it cites.

\section{Answers to the Research Questions}
\label{sec:rq_answers}

\paragraph{RQ1 (capability).}
\emph{Yes: the pipeline suppresses false positives while raising, not merely maintaining, detection performance.} On the discriminating Svanstr\"om benchmark the full production pipeline lifts drone $F1$ from $0.742$ (bare detectors) to $0.946$, with recall \emph{rising} from $0.948$ to $0.991$ and precision from $0.609$ to $0.905$ (Table~\ref{tab:ablation_svanstrom}). On the out-of-distribution confuser corpus, where the bare detector fires on $30.4\%$ of frames, the per-frame filter reduces fire to $1.4\%$; the reject-class \texttt{robust8} ablation drives it to $0.11\%$ (three frames in $2{,}633$, Table~\ref{tab:ablation_confusers}), the suppression the shipped no-reject stack trades for recall. The thermal-confuser surface, formerly the pipeline's weakest front, is largely closed by the thermal-native IR filter: the shipped no-reject stack cuts ${\sim}91\%$ of thermal-confuser fire ($29.4\% \to 2.8\%$, Table~\ref{tab:ablation_confusers}), and airplanes --- once the resistant class --- no longer dominate the residual (the honest held-out IR\_confusers val/test figure is $94\%$ removed, Section~\ref{sec:ir_xmodal_verifier}). On the saturated Anti-UAV control the pipeline does no harm ($F1$ $0.973 \to 0.984$, Table~\ref{tab:ablation_antiuav}); on ordinary, confuser-free footage the false-alarm problem was already small at the detector ($41$ false positives across $4{,}000$ Anti-UAV frames), which is precisely why the thesis locates the problem in confuser scenes and solves it there.
% [source: thesis_eval/results/tier1_results.json (svanstrom/antiuav B_pipeline, rgb_confuser C_confuser)]

\paragraph{RQ2 (attribution).}
\emph{Each stage earns its place, and they are complementary rather than redundant.} The router and the filter address different false positives. For the shipped no-reject router the per-frame filter is the workhorse: on Svanstr\"om it removes $83\%$ of bare false positives ($2{,}019 \to 337$) while the no-reject router, routing every frame, removes almost none on its own; on the confuser corpus the filter is again the decisive stage ($835 \to 39$ FP detections, where the patch-CNN predecessor managed $835 \to 282$, ${\sim}7\times$ fewer). The reject-class \texttt{robust8} ablation supplies the complementary mechanism, vetoing whole untrustworthy frames ($82\%$ of bare FPs at the router alone), the net the shipped stack trades for recall. Composition order is a recall/precision dial: the shipped filter-then-classify keeps the router's recall ($R=0.991$, $F1=0.946$) and classify-then-filter trades a little recall for precision ($F1=0.931$), within $1.5$~pp, so the choice is operational, not structural. Temporal aggregation converts per-frame decisions into alert events; the segment-level evaluation on the fully-OOD video surface (Table~\ref{tab:temporal_production}) is where the costs show and where a design lesson lands. Per-frame routing cuts confuser fire $71$--$77\%$ for a $10$--$17$~pp window-F1 cost, but a \emph{per-frame} filter veto, with thresholds tuned on benchmark stills, over-vetoes OOD drones; a threshold sweep over the cached probabilities shows the cost is structural rather than tunable (true-drone probabilities smear across the operating range), whereas the predecessor's \emph{alert-gated} veto on the same surface gained F1. The gate position, not the gate, is the variable; the deployment accordingly routes per-frame and verifies at the alert gate on OOD video. The runtime accounting closes the argument: the production router costs $0.095$~ms per frame and the filter $1.3$--$2.1$~ms per detection ($404\times$ and $37$--$72\times$ cheaper than their predecessors respectively, $1$--$4\%$ total overhead), which is why every stage can run on every frame and the predecessor's alert-gating compromise disappears (Table~\ref{tab:speed}).
% [source: thesis_eval/results/tier1_results.json; ledger=cascade-segment-recovers,robust6-speed-feature-efficiency,v5-ship-per-frame]

\paragraph{RQ3 (dual-modality value).}
\emph{Yes: the modality ranking reverses across surfaces, and routing pays on both sides of the reversal.} On Svanstr\"om the thermal detector dominates the visible one ($F1$ $0.940$ vs $0.607$); on Anti-UAV the visible detector dominates ($0.985$ vs $0.961$); the routed pipeline matches or beats the better single modality on each (Table~\ref{tab:rq3}). The accounting nuance matters: the thermal channel's value concentrates exactly on frames where the visible channel fails, so each modality is scored against its own ground truth and never unioned; under a union rule the same detections would mis-attribute the contribution by double-digit F1 (Section~\ref{sec:scoring_audit}).

\paragraph{The methodological thread.}
Both contributions behind the components delivered measurable value. The \emph{Model MRI}'s statistics-before-training discipline trained the filter family directly from measured drone/confuser separability (linear separability $\approx$95\% RGB / $0.981$ IR; drone detections as positives, confuser detections as negatives), and selected the router's eight features by leakage analysis. Re-running it on its own shipped corpus corrected two figures in an earlier report (Section~\ref{sec:mri_findings}). The \emph{HITL data engine} carried the IR detector from $F1$ $0.503$ to $0.967$ across six revisions on a fixed test split; the one revision that bypassed the review loop (V5) cost $12.7$~points of precision and is reported as the protocol's clearest negative result: the dataset state, not the model state, is what regressed (Section~\ref{sec:ir_evolution}).
% [source: ledger=ir-version-progression,mri-v5-report-regen]

\paragraph{The findings.}
The grayscale result stands as measured: a thermal-only-trained detector, run zero-shot on grayscale-converted RGB, still detects drones when the visible channel fails, landing within $2.7$~pp F1 of the dedicated RGB detector at matched configuration ($0.580$ vs $0.607$), with higher precision and lower recall, and tying it on the hardest bird-cluttered clip. Its raw-RGB control collapses on both surfaces ($F1=0.187$ on Svanstr\"om, $0.295$ on video), isolating the single-channel conversion as the load-bearing step (Section~\ref{sec:grayscale}). Its practical payoff is a recall-preserving fallback when thermal hardware is unavailable.

\section{Production Stack}
\label{sec:production_stack}

The stack, as validated in this thesis: RGB detector \texttt{ft4} (the SelCom-plus-confuser fine-tune; the only confuser-injection recipe that passed every drone-recall regression gate); IR detector \texttt{v3b}; trust router \texttt{robust8-nr} (the eight free features selected by the leakage statistic, with the \texttt{reject} class dropped so the router always routes and the filter owns false-positive rejection), composed in the filter-then-router order (\texttt{filt$\to$clf}, which keeps recall at the router's level); \texttt{robust8} with reject is retained as an ablation and is the recommended choice on confuser-rich paired streams; per-frame filters \texttt{mlp\_v5} (RGB, the \texttt{v4} bird-split build, @0.25) and the thermal-native IR filter (\texttt{mlp\_aligned\_thermalonly}, CBAM-trained, @0.05); an $N$-of-$M$ temporal smoother above them; inference at \texttt{imgsz=1280} on Svanstr\"om-like and SelCom-like content (960 is the measured deployment optimum for the SelCom camera) and 640 elsewhere; trust-aware scoring as the reporting rule. One validated \emph{operating mode} extends it: on recall-starved OOD RGB surfaces, the RGB confidence floor can drop to $0.05$--$0.10$ with the filter owning precision. The floors existed to control false positives, and that job has moved downstream ($+10$~pp F1 on SelCom at unchanged confuser safety; Section~\ref{sec:lowconf_mode}).

Two carve-outs accompany it, each diagnosed rather than hidden. The earlier \texttt{rgb\_dataset} coverage gap --- the shipped \texttt{mlp\_v5} cost $11$~pp F1 there --- is now \emph{closed}: the \texttt{v4} bird-split build lifts \texttt{rgb\_dataset\_test} F1 from $0.809$ to $0.922$ (recall $0.691 \to 0.887$), so the patch filter is no longer needed as a fallback on that surface (Section~\ref{sec:verifier_results}). \emph{(i)} the shipped no-reject \texttt{robust8-nr} trades confuser suppression for recall: RGB-confuser frame fire rises to $\sim$1.4\% and real-video-confuser fire $\sim$3$\times$ relative to the reject-class \texttt{robust8}, which is retained as the recommended router where confuser-rich paired video dominates (Section~\ref{sec:classifier_results}). \emph{(ii)} The thermal-native IR filter is validated per-frame in the evaluation engine; its operator-GUI wiring is an open engineering item at the time of writing, as is Jetson-class edge latency.

\section{Future Work}

Four directions. A \emph{track-based classifier} consuming the temporal smoother's per-track state (velocity, persistence, trajectory curvature) as a fourth evidence channel, the most promising single direction, and the one aimed at the airplane class that appearance-based stages handle least well. \emph{Further hardening the airplane class}: the thermal-native filter already cut the airplane-dominated thermal-confuser fire $29.4\% \to 2.8\%$ (cache) / $94\%$ (held-out), and additional OOD airplane crops as confuser-class training data would press the residual lower. \emph{Completing the engineering}: GUI wiring of the aligned filter and TensorRT/ONNX edge benchmarking. And \emph{extending the protocol}: the open-world \texttt{fusion\_no\_fn\_v1.1} router remains selectable where missed drones are operationally cheaper than false alarms, and the dual-classifier extension (Section~\ref{sec:classifier_results}) is validated if a deployment needs the missing-modality regimes as first-class citizens.
