\chapter{Introduction}
\label{ch:introduction}

\section{Background}

Consumer unmanned aerial vehicles (UAVs) are inexpensive, widely available, and increasingly implicated in unauthorised surveillance, perimeter incursion, and disruption of civil airspace operations~\cite{shi2018counteruas,taha2019drone,samaras2019deep}. Protecting a fixed site against them imposes two simultaneous requirements on any sensing system: \emph{high recall}, because a missed drone is an unrecoverable failure of the system's purpose, and a \emph{low false-alarm rate}, because frequent alerts for birds and passing aircraft risk desensitising the operator the system is meant to serve. The Counter-UAS (C-UAS) sensing surveys~\cite{shi2018counteruas,taha2019drone,samaras2019deep} document the limitations of each sensing modality: radar suffers on the small cross-section of quadcopter-class drones; radio-frequency detection requires an active command link and is defeated by autonomous pre-programmed flight, and more recently by fiber-optic-guided drones, fielded at scale since 2024, whose tethered control link emits no RF signature at all~\cite{csis2025dronewar}; acoustic arrays degrade under wind, traffic, and aircraft noise. Visual detection with deep neural detectors is passive, low-cost, and context-rich, and is the modality adopted in this work.

Modern single-stage detectors are genuinely strong on this task. The YOLO family~\cite{redmon2016yolo}, as packaged in the Ultralytics framework~\cite{ultralytics2024}, delivers high drone-class precision and recall at real-time rates. The production RGB detector of this thesis reaches $P=0.989$, $R=0.982$ on the Anti-UAV benchmark~\cite{jiang2021antiuav}, raising only 41 false positives across 4{,}000 even-spread paired frames, and hallucinates on $2.8\%$ of its composite RGB test corpus.\footnote{Anti-UAV frames appear in the detector's training corpus, so this is an in-distribution sanity floor rather than a generalisation claim (Section~\ref{sec:lit_drone}); evaluated at \texttt{imgsz=640}, $n=4{,}000$.} On \emph{ordinary} footage --- sky, buildings, traffic, vegetation --- the false-alarm problem is, for the detectors trained here, largely solved at the detector.
% [source: thesis_eval/results/tier1_results.json (antiuav A ft4/rgb); run=thesis_eval/pipeline_eval_unified.py; config=tier1 antiuav @640 n=4000]
% [source: eval=v5_rgbds (bare halluc 0.028); config=rgb_dataset_iou_640]

\section{Problem Statement}
\label{sec:problem}

The false-alarm problem is not generic; on much footage the detector never fires falsely at all. It is specific to scenes containing the three aerial confuser categories (birds, fixed-wing aircraft, and helicopters). On the \texttt{rgb\_confusers\_merged} test split --- a 2{,}633-image out-of-distribution confuser corpus assembled for this thesis, two-thirds Svanstr\"om-sourced and fully unseen by the production detector's training (Section~\ref{sec:ds_confusers}) --- the same detector family fires on $30.4\%$ of frames, rising to as much as $94\%$ on bird-only Svanstr\"om frames at that benchmark's canonical \texttt{imgsz=1280} (the per-category breakdown is in Table~\ref{tab:rgb_comparison}; Figure~\ref{fig:confuser_fp_examples} shows representative hallucinations). At surveillance ranges all four object classes present as small, low-texture silhouettes against sky, and the appearance gap that separates them is too narrow for the detector to exploit \emph{at its decision layer} without collateral damage: as the next paragraph shows, retraining against confusers trades away drone recall. The separation does, however, demonstrably exist deeper in the detector, in its own intermediate feature space (Section~\ref{sec:model_mri}), and exploiting it \emph{there} instead of at the detection head is precisely the design of this thesis.
% [source: thesis_eval/results/tier1_results.json (rgb_confuser C bare); thesis_eval/results/notes_round1_results.json (rgb_confuser CAT); config=tier1 rgb_confuser @640 n=2633]
% [source: ledger=retrainedv2-recall-collapse; eval=rgb_svan_baseline; cache=eval/results/_failure_diagnosis/svanstrom_1280_by_category.csv; config=svan_iop_1280]

The obvious mitigation, retraining the detector with confuser images as hard negatives, has diminishing and \emph{class-specific} returns: \texttt{hardneg\_v3more}, trained with Svanstr\"om's confuser splits as explicit negatives, lowers helicopter and airplane fire but leaves bird fire essentially unchanged (Table~\ref{tab:rgb_comparison}). A more aggressive retraining (\texttt{retrained\_v2}) does suppress bird fire, but collapses drone recall on Svanstr\"om from $R=0.961$ to $R=0.306$ at the same precision. The collapse is out-of-distribution-specific rather than a general defect: on its own in-domain test split the same weights score $F1=0.949$, \emph{above} the baseline's $0.942$, so in-domain selection would pick exactly the detector that fails on deployment-like small-drone footage. Detector recall, once lost, cannot be recovered downstream: no later stage adds detections the detector never proposed.
% [source: ledger=retrainedv2-recall-collapse; eval=rgb_svan_hardneg,rgb_svan_retrainedv2; cache=eval/results/_failure_diagnosis/svanstrom_1280_by_category.csv; config=svan_iop_1280]
% [source: ledger=rgb-collapse-ood-specific; eval=rgb_rgbds_retrainedv2 (0.9494 in-domain > baseline 0.942); config=rgb_dataset_iou_640]

Two further gaps separate benchmark performance from field readiness. \emph{Resolution}: small or distant drones fall below the resolvable floor at standard inference sizes and disappear from the detector entirely (Section~\ref{sec:resolution_arch}); the per-size breakdown of Section~\ref{sec:per_size} makes this concrete (on Svanstr\"om, where the median drone ground-truth $\sqrt{\text{area}}$ is 29.8~px in the native $640{\times}512$ frame, bare RGB recall falls from $0.90$--$0.96$ in the 16--64~px buckets to $0.63$ below 16~px). \emph{Modality complementarity}: the visible-light channel degrades at night and under glare and low contrast, exactly where the thermal channel is strongest, so the system is dual-modality from the start. Which modality dominates \emph{reverses} with the scene (Section~\ref{sec:rgb_results}): the thermal detector leads on Svanstr\"om (own-GT F1 $0.940$ vs $0.607$), the visible detector leads on Anti-UAV ($0.985$ vs $0.961$), and the routed system matches the stronger one on each (Table~\ref{tab:rq3}).
% [source: thesis_eval/results/notes_round1_results.json (svanstrom SZ, M); appendix tab Svanström native 640x512]

Any discrimination the detector cannot learn without costing recall must therefore be moved \emph{off} the detector. This thesis relocates confuser discrimination, modality arbitration, and false-positive filtering into a downstream dual-modality system: a trust classifier that routes between an RGB and a thermal detector per frame, a per-detection \emph{MLP confuser filter} that re-reads the detector's own internal features and metadata --- already computed for each detection --- to remove confuser false positives from the survivors (Section~\ref{sec:distill_verifier}); and a temporal aggregator that turns per-frame decisions into alerts. The base detector is left tuned purely for recall.

One evaluation surface comes from a deployment partner: a fixed perimeter-CCTV installation (SelCom) whose 1080p stream renders drones at a median $\sqrt{\text{area}}$ of $\approx$37~px in the native $1920{\times}1080$ frame amid urban clutter. The system is not deployed there; the partner provided RGB footage on which the RGB branch was fine-tuned and evaluated (the \texttt{ft4} fine-tune lifted SelCom validation F1 from the baseline's $0.145$ to $0.591$ while leaving the composite in-domain test corpus essentially unchanged, $0.942 \to 0.929$). The field gaps above were observed across \emph{all} of the evaluation surfaces of this thesis, with the CCTV footage contributing the small-drone-in-clutter case. Two engineering constraints hold throughout: every component must run interactively on modest hardware with no specialised accelerators beyond a consumer GPU (the stage-cost budget of Section~\ref{sec:pipeline_speed} quantifies this), and every reported number must correspond to a configuration the system can actually ship.
% [source: dataset=selcom_cctv; file=SelCom CCTV manifest (median 36.8px native)]
% [source: eval=v5_selcom_bare (ft4 0.5911), selcom baseline 0.145 (eval/results _selcom cmp); eval=rgb_rgbds_retrainedv2 extra (baseline 0.942, ft4 0.929)]

\section{Research Questions}
\label{sec:rqs}

The thesis is organised around three research questions; answers are mapped to evidence in Section~\ref{sec:rq_answers}.

\begin{description}
    \item[RQ1 (capability)] Retraining the base detector is the cheap first answer to false alarms, and this work tried it first: hard-negative retraining either leaves the dominant confuser class untouched or collapses small-drone recall (Section~\ref{sec:problem}). \emph{Where retraining fails}, can a layered detection pipeline maintain or improve drone-detection performance while suppressing false positives, on both ordinary and confuser-rich scenes?
    \item[RQ2 (attribution)] What does each stage of the pipeline contribute? Measured by ablation: the trust classifier, the \emph{confuser filter} (the per-detection MLP defined above), their composition order, and temporal evidence aggregation, each isolated on the same frames and with the runtime cost of each stage accounted for.
    \item[RQ3 (dual-modality value)] Does per-frame routing between two modalities outperform either modality alone? The scoring protocol this question requires is defined in Section~\ref{sec:metrics}.
\end{description}

A methodological discipline supports the downstream stages: \emph{statistics before training}. For the trust classifiers and the confuser filters (and only for those; the detectors were trained conventionally), a model-agnostic feature-space instrument (the Model MRI, Section~\ref{sec:model_mri}) measured what the detectors' features can separate \emph{before} any training run, and a leakage-ratio statistic, applied to the candidate routing features, excluded scene-fingerprint features from the trust classifier. Separately, a human-in-the-loop review process co-develops the IR detector with its training corpus.

\section{Contributions}

This thesis makes four contributions, and reports one finding that emerged from them.

\begin{enumerate}
    \item A \emph{deployable dual-modality drone-detection system with an operator GUI}. An RGB and a thermal-infrared YOLO detector are fused by an XGBoost trust classifier that makes a per-frame four-way modality decision (\texttt{reject} / \texttt{trust\_rgb} / \texttt{trust\_ir} / \texttt{trust\_both}); the trusted modality's detections are then filtered by the per-detection MLP confuser filter (Section~\ref{sec:distill_verifier}), and a temporal smoother converts per-frame decisions into alerts. Because earlier project numbers had mixed inference sizes, matching rules, and splits and were not mutually comparable, the system is held to one declared evaluation standard (Section~\ref{sec:eval_protocol}); under it the full pipeline lifts Svanstr\"om paired drone $F1$ from $0.742$ to $0.946$ while \emph{raising} recall, holds the saturated Anti-UAV control to no harm, and cuts out-of-distribution confuser fire by more than an order of magnitude, every stage running on every frame within budget. Chapter~\ref{ch:hitl} reports the per-stage ablation and the full numbers, and Appendix~\ref{app:provenance} maps each headline figure to its results file and replay command in the \texttt{runs/} directory.
    % [source: thesis_eval/results/tier1_results.json (svanstrom B, rgb_confuser C, antiuav B); run=thesis_eval/pipeline_eval_unified.py]
    % [source: ledger=robust6-speed-feature-efficiency,v5-ship-per-frame; run=eval/bench_speed.py]
    \item The \emph{Model MRI}, a statistics-before-training instrument for detector feature spaces. It hooks a YOLO detector's pyramid features, reads a 517-D per-detection embedding (internal activations plus detection metadata), and answers design questions with classical statistics before any model is trained: linear-discriminant separability, per-feature ANOVA, single-feature AUROC, a leakage ratio that exposes scene-fingerprint features (applied in this work to select the trust classifier's routing features), and per-modality standardisation diagnostics. Its measurements then train the filters directly: where the feature space is separable, the MRI trains a lightweight MLP confuser filter on the detector's own internal-feature embeddings (positives = drone detections, negatives = confuser detections). Drone/confuser linear separability measured 0.952 (RGB) and 0.981 (IR) on the detectors' own cached feature corpora; these are in-corpus diagnoses that license a training attempt, and the filter's out-of-distribution evals (Chapter~\ref{ch:hitl}) are what validate it.
    % [source: ledger=v5-lda-separability,ir-features-separable; run=mri/cli.py]
    \item A \emph{human-in-the-loop data engine}: a purpose-built label-reviewer GUI and a disciplined dataset--model co-evolution loop, reported as a six-revision case study of the IR detector. Every revision, first to last, is scored on the same fixed held-out test split of the final IR corpus ($F1\;0.503 \to 0.967$, $n=9{,}612$), so the trajectory is one comparable series rather than each model on its own contemporary split. The loop exists because clean thermal drone data is scarce: public drone datasets are difficult to obtain in RGB, far more so in thermal or paired form, and the available material needed frame-level review. The trace includes the one revision where the protocol was violated (V5: a bulk positive ingestion bypassed review and cost 12.7 points of precision), and the recovery was itself a product of the loop: the same review process that the ingestion bypassed is what located and corrected the regression.
    % [source: ledger=ir-version-progression; eval=ir_final_* (all on config ir_final_640, same split)]
    \item A set of \emph{curated data artifacts}, the product of the manual review work behind the system. On the corpus side: the 129{,}130-frame thermal corpus, frame-level reviewed across the six dataset revisions that co-evolved with the IR detector (Section~\ref{sec:hitl_method}); the 27{,}024-frame RGB confuser hard-negative corpus with sequence-disjoint splits (Section~\ref{sec:ds_confusers}); and the paired Svanstr\"om conversion, 28{,}710 aligned visible/thermal frame pairs converted from the dataset's MATLAB \texttt{groundTruth} annotations to YOLO format. On the benchmark side: a 19-clip real-video diagnostic, 2{,}609 frames with 1{,}234 ground-truth drone instances annotated frame by frame, the only evaluation surface in this thesis that is fully out-of-distribution for \emph{every} pipeline component (Section~\ref{sec:ds_youtube}). The source datasets' licences do not permit redistributing the corpora, so they are claimed as reviewed artifacts rather than released data; the real-video benchmark, by contrast, is largely reproducible outside this project: its labels are publishable, and 17 of the 19 clips carry recoverable source identifiers in Appendix~\ref{app:datasets} (two confuser compilations' identifiers were not preserved).
    % [source: tab:ds_ir_components total 129,130; rgb_confusers_merged 27,024; svanstrom_paired 28,710; sec:ds_youtube 19 clips / 2,609 frames / 1,234 GT instances]
\end{enumerate}

One finding emerged that was not designed: a \emph{cross-modal transfer} result. The thermal-trained IR detector, run zero-shot on grayscale-converted RGB, performs within a few points of the dedicated RGB detector on the same frames ($F1$ $0.580$ vs $0.607$ on the Svanstr\"om visible channel, $n=4{,}000$, matched configuration) and ties it on the hardest bird-cluttered video clip (\texttt{flock\_of\_seagulls\_attack\_drone\_beach}, $F1\;0.837$ vs $0.840$). (The near-linear separability of drone from confuser inside the detector's own internal features, which lets the 517-D MLP outperform the MobileNetV3 patch classifier, is the premise behind the Model MRI contribution above and is shown in Section~\ref{sec:verifier_results}.)
% [source: thesis_eval/results/tier1_results.json (D table); ledger=ir-grayscale-fallback (clip + 0.837/0.840); eval=patch_catch_v2_svan,v5_svan_mlp/patch]

Evaluation is held to a single declared standard: one canonical configuration per dataset (inference size, matching rule, frame budget), per-modality trust-aware scoring, frame counts printed with every number, and bootstrap confidence intervals on headline results (Section~\ref{sec:eval_protocol}). A reproducibility layer (per-run JSON manifests, weight-hash-tagged inference caches, and a unified detection cache that makes every downstream ablation a deterministic replay) ties every reported number to the command that produced it, and Appendix~\ref{app:provenance} gives the reader the concrete map: headline number $\to$ canonical results file in \texttt{runs/} $\to$ replay command $\to$ automated audit check. Each evaluated model additionally ships a co-located \texttt{*.model\_card.yaml} naming its train and test surfaces, so a reader can open the provenance of any weight directly (Appendix~\ref{app:models}).

\section{Ethical Considerations and Dual-Use}
\label{sec:ethics}

Counter-UAS sensing systems are dual-use technology. The same passive visual-detection capability that flags a hostile drone over critical infrastructure can, in a different deployment, contribute to mass aerial surveillance or armed-response targeting. This thesis develops the detection component only; it neither integrates with effectors (jammers, kinetic intercept, take-down systems) nor classifies operator behaviour. The training corpora contain no personally identifying information, no identifiable third-party imagery beyond what is publicly available in the cited datasets and YouTube clips, and no acoustic or RF data that could enable operator identification. The deployment scope assumed throughout is protection of fixed sites (perimeter monitoring, airport approaches, industrial facilities) where a detected drone is reported to a human operator who decides on response. The reproducibility layer (manifests, hash-tagged caches, and the number-provenance map of Appendix~\ref{app:provenance}) is intended in part to make any future adaptation auditable: an auditor can re-run the replay commands against the published caches and verify that the reported behaviour is the shipped behaviour.

\section{Thesis Outline}
\label{sec:outline}

Chapter~\ref{ch:literature} reviews the literature on real-time object detection, drone detection and the confuser problem, thermal imaging, multi-modal fusion, detection scoring, cross-modal transfer, cascaded rejection, and the two methodological literatures this work builds on: model-assisted annotation and feature-space analysis. Chapter~\ref{ch:methodology} establishes the empirical infrastructure: the datasets and their roles, the evaluation protocol and its justifications, the system architecture, each component's design and training, the Model MRI, and the human-in-the-loop method. Chapter~\ref{ch:hitl} reports the empirical evaluation: the full-pipeline ablation that answers RQ1--RQ3, the temporal segment-level evidence, the per-component deep dives (including the IR detector's six-revision evolution), the Model MRI findings, the grayscale cross-modal finding, and threats to validity. Chapter~\ref{ch:conclusion} answers the research questions, states the production stack with its honest carve-outs, and lists future work. Appendix~\ref{app:datasets} details the datasets; Appendix~\ref{app:models} the models evaluated; Appendix~\ref{app:glossary} the glossary.
